Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-10-11 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Tuned language.

The results in the previous section revealed significant differences of the glycerol backbone and choline headgroup
order parameters between different molecular dynamics simulation models.
However, it is not straightforward to conclude which kind of structural differences (if any)
between the models the results indicate, because the mapping from the order parameters to the 
structure is not unique. In this section we demonstrate that 1) the differences in order parameters
indicate significantly different structural sampling strongly correlating with the dihedral angles of the related bonds,
and that 2) the comparison between experimental and simulated order parameters can be used to exclude
nonrealistic structural samping in molecular dynamics simulations. The demonstration is done for 
the dihedral angles defined by the g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) segments in the glycerol backbone and 
the N-$\beta$-$\alpha$-O segments in the headgroup. These dihedrals were chosen for demonstration, because 
significant differences between the models are observed around these segments in Fig.~\ref{HGorderparameters2}.
We note that performing a similar comparison through all the dihedrals in all the 13 models would probably give highly useful
information on how to improve the accuracy of the models yet this is beyond the scope of the current report.

The dihedral angle distributions for the  g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral calculated from different models are
shown in Fig.~\ref{dihDISTS}. The distribution is qualitatively different for the Berger-POPC-07 model, showing a maximum in 
the gauche$^+$-conformation (60$^o$) compared to all the other models showing a maximum in the anti-conformation (180$^o$).
The distributions in all the other models have the same general features, the main difference being that the
fraction of configurations in the gauche$^-$-conformation (-60$^o$) is zero for the MacRog, detectable for the CHARMM36 and
equally large to the gauche$^+$ fraction in GAFFlipid. From the results we conclude that most likely the wrongly sampled
dihedral angle for the g$_2$-g$_1$ bond explains the significant discrepancy to the experimental order parameters
for the g$_1$ segment in the Berger-POPC-07 model (Fig.~\ref{HGorderparameters2}). 
In conclusion, models preferring the anti conformation for this dihedral give more realistic order parameters and
this is in agreement with previous crystal structure and $^1$H NMR studies~\cite{hauser80,hauser81,hauser81b,hauser88,pascher92,marsh06}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../Fig/g1-g2_Cdihs2.eps}
  \caption{\label{dihDISTS}
    Dihedral angle distributions for g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral from different models (POPC bilayer in full hydration).
      } 
\end{figure}

The dihedral angle distribution for the  N-$\beta$-$\alpha$-O dihedral calculated from the same four models is 
shown in Fig.~\ref{dihDISTS2}. Also for this dihedral there are significant differences in the gauche--anti fractions.
The gauche conformations are dominant in the CHARMM36, in MacRog there are only anti conformations present,
and in the Berger-POPC-07 and GAFFlipid gauche and anti conformations have equal probabilities. 
On the other hand, comparison of $\alpha$ and $\beta$ order parameters in Fig.~\ref{HGorderparameters2}
reveals that for these carbons the CHARMM36 is closest to the experimental results and it is also the only model that has the correct
sign (negative) for the $\beta$ order parameter. This result is again in agreement with previous 
crystal structure, $^1$H NMR and Raman spectroscopy studies~\cite{hauser80,hauser81,hauser81b,akutsu81b} which suggest that
this dihedral is in the gauche conformation in the absence of ions.

@@

The used examples show that the glycerol backbone and headgroup order parameters reflect the atomistic resolution structure
and that the comparison with experiments allows the assessment of the quality of the suggested structure. We were able to pinpoint
specific problems in the structures in different models and suggest potential improvement strategies.
If the improved atomistic
molecular dynamics simulation model reproduced the order parameters and other experimental observables 
(like $^{31}$P chemical shift anisotropy~\cite{chowdhary13} and $^{31}$P-$^{13}$C dipolar couplings~\cite{prakash10})
with experimental accuracy, it would give an interpretation for the atomistic resolution structure of the glycerol backbone and 
choline~\cite{seelig77b,skarjune79,jacobs80,davis83,akutsu91,hong95b,semchyschyn04}. The research along these lines is left, however,
for future studies.

>>>>

The results in the previous section revealed significant differences of the glycerol backbone and choline headgroup
order parameters between different molecular dynamics simulation models.
However, it is not straightforward to conclude which kind of structural differences (if any)
between the models the results indicate, because the mapping from the order parameters to the 
structure is not unique. In this section we demonstrate that 1) the differences in order parameters
indicate significantly different structural sampling, which is strongly correlated with the dihedral angles of the related bonds;
and that 2) the comparison between experimental and simulated order parameters can be used to exclude
nonrealistic structural sampling in molecular dynamics simulations. The demonstration is done for 
the dihedral angles defined by the g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) segments in the glycerol backbone and 
the N-$\beta$-$\alpha$-O segments in the headgroup. These dihedrals were chosen for demonstration, because 
significant differences between the models are observed around these segments in Fig.~\ref{HGorderparameters2}.
We note that performing a similar comparison through all the dihedrals in all the 13 models would probably give highly useful
information on how to improve the accuracy of the models; yet this is beyond the scope of the current report.

The dihedral angle distributions for the  g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral calculated from different models are
shown in Fig.~\ref{dihDISTS}. The distribution is qualitatively different for the Berger-POPC-07 model, showing a maximum in 
the gauche$^+$-conformation (60$^o$) compared to all the other models showing a maximum in the anti-conformation (180$^o$).
The distributions in all the other models have the same general features, the main difference being that the
fraction of configurations in the gauche$^-$-conformation (-60$^o$) is zero for the MacRog, detectable for the CHARMM36 and
equally large to the gauche$^+$ fraction in GAFFlipid. From the results we conclude that most likely the wrongly sampled
dihedral angle for the g$_2$-g$_1$ bond explains the significant discrepancy to the experimental order parameters
for the g$_1$ segment in the Berger-POPC-07 model (Fig.~\ref{HGorderparameters2}). 
In conclusion, models preferring the anti conformation for this dihedral give more realistic order parameters;
this is in agreement with previous crystal structure and $^1$H NMR studies~\cite{hauser80,hauser81,hauser81b,hauser88,pascher92,marsh06}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../Fig/g1-g2_Cdihs2.eps}
  \caption{\label{dihDISTS}
    Dihedral angle distributions for g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral from different models (POPC bilayer in full hydration).
      } 
\end{figure}

The dihedral angle distribution for the  N-$\beta$-$\alpha$-O dihedral calculated from the same four models is 
shown in Fig.~\ref{dihDISTS2}. Also for this dihedral there are significant differences in the gauche--anti fractions.
The gauche conformations are dominant in CHARMM36, in MacRog there are only anti conformations present,
and in Berger-POPC-07 and GAFFlipid the gauche and anti conformations have equal probabilities. 
On the other hand, comparison of $\alpha$ and $\beta$ order parameters in Fig.~\ref{HGorderparameters2}
reveals that for these carbons the CHARMM36 is closest to the experimental results and it is also the only model that has the correct
sign (negative) for the $\beta$ order parameter. This result is again in agreement with previous 
crystal structure, $^1$H NMR, and Raman spectroscopy studies~\cite{hauser80,hauser81,hauser81b,akutsu81b}, which suggest that
this dihedral is in the gauche conformation in the absence of ions.

@@

These examples show that the glycerol backbone and headgroup order parameters reflect the atomistic resolution structure
and that the comparison with experiments allows the assessment of the quality of the suggested structure. We were able to pinpoint
specific problems in the structures in different models and suggest potential improvement strategies.
If an improved atomistic
molecular dynamics simulation model would reproduce the order parameters and other experimental observables 
(like $^{31}$P chemical shift anisotropy~\cite{chowdhary13} and $^{31}$P-$^{13}$C dipolar couplings~\cite{prakash10})
within experimental accuracy, it would give an interpretation for the atomistic resolution structure of the glycerol backbone and 
choline~\cite{seelig77b,skarjune79,jacobs80,davis83,akutsu91,hong95b,semchyschyn04}. The research along these lines is left, however,
for future studies.
